---
title: "Final_Numbers"
author: "Annie Kellner"
date: "2025-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r load-libraries, message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)
library(unmarked)
library(MuMIn)
library(sf)
library(tmap)
library(tmaptools)
library(viridisLite)

```

```{r load-data}

umf <- readRDS("~/Repos/MGS/Data/UMFs/umf_FINAL_20251003.Rds")

df <- readRDS("~/Repos/MGS/Data/finalDF_shpHabitat_camsDeleted_20251003.Rds")

```


### DETECTION AIC


## Model 1: Intercept-Only

```{r intercept-only, warning=FALSE}

m1 <- occu(~ 1 ~ 1,
           data = umf,
           linkPsi = "logit")

```

## Model 2: Linear Date

```{r D-linear, warning=FALSE}

Dm2 <- occu(~ scaled_recentered ~ 1,
           data = umf,
           linkPsi = "logit")

```


## Model 3: Quadratic Date

```{r D-quadratic, warning=FALSE}

Dm3 <- occu(~ scaled_recentered + scaled_recentered_squared ~ 1,
           data = umf,
           linkPsi = "logit")

```


## Model 4: Season Only

```{r D-Season, warning=FALSE}

Dm4 <- occu(~ season ~ 1,
           data = umf,
           linkPsi = "logit")

```

## Model 5: Linear Date + Season

```{r D-linear-season, warning=FALSE}

Dm5 <- occu(~ scaled_recentered + season ~ 1,
           data = umf,
           linkPsi = "logit")

```


## Quadratic + Season 

```{r D-quadratic-season, warning=FALSE}

Dm6 <- occu(~ scaled_recentered + scaled_recentered_squared + season ~ 1,
           data = umf,
           linkPsi = "logit")

```

*AIC TABLE - DETECTION*

```{r D-AIC}

model_list <- list(
  "Intercept-Only" = m1,
  "Linear Date" = Dm2,
  "Quadratic Date" = Dm3,
  "Season" = Dm4,
  "Linear Date + Season" = Dm5,
  "Quadratic Date + Season" = Dm6
)

# Calculate AICc for each model

aicc_values <- sapply(model_list, function(x) {
  AICc(x)
})

# Create comparison table

model_comparison <- data.frame(
  Model = names(model_list),
  AICc = aicc_values,
  Delta_AICc = aicc_values - min(aicc_values)
)

# Sort by AICc

model_comparison <- model_comparison[order(model_comparison$AICc),]

model_comparison

```


### OCCUPANCY AIC

# Model 1 is Dm6 from detection AIC (Detection Covs with Intercept on Occupancy)

## Model 2: all Habitat

```{r all-habitat}

Om2 <- occu(~scaled_recentered + scaled_recentered_squared + season ~Habitat,
            data = umf,
            linkPsi = "logit")

```

## Model 3: Joshua Tree

```{r}

Om3 <- occu(~scaled_recentered + scaled_recentered_squared + season ~JoshTree,
            data = umf,
            linkPsi = "logit")

```

## Model 4: Desert Scrub

```{r scrub, warning=FALSE}

Om4 <- occu(~scaled_recentered + scaled_recentered_squared + season ~DesertScrub,
            data = umf,
            linkPsi = "logit")

```


## Model 5: Distance to Stream


```{r Distance, warning=FALSE}

Om5 <- occu(~scaled_recentered + scaled_recentered_squared + season ~Dist_Scaled,
           data = umf,
           linkPsi = "logit")

```


## Model 6: Habitat + Distance to Stream

```{r Habitat-Distance, warning=FALSE}

Om6 <- occu(~ scaled_recentered + scaled_recentered_squared ~ Habitat + Dist_Scaled,
           data = umf, 
           linkPsi = "logit")


```


## Model 7: Joshua Tree + Distance to Stream 

```{r JTree-Distance, warning=FALSE}

Om7 <- occu(~ scaled_recentered + scaled_recentered_squared ~ Dist_Scaled + JoshTree,
           data = umf,
           linkPsi = "logit")

```


## Model 8: Desert Scrub + Distance to Stream

```{r date-season-distance, warning=FALSE}

Om8 <- occu(~ scaled_recentered + scaled_recentered_squared + season ~ Dist_Scaled + DesertScrub,
           data = umf,
           linkPsi = "logit")

```


## AIC Analysis

```{r }

model_list <- list(
  "Detection Covs Only" = Dm6, # from detection model set
  "All Habitat" = Om2,
  "Joshua Tree" = Om3,
  "Desert Scrub" = Om4,
  "Distance to Stream" = Om5,
  "All Habitat + Distance to Stream" = Om6,
  "Joshua Tree + Distance to Stream" = Om7,
  "Desert Scrub + Distance to Stream" = Om8
)

# Calculate AICc for each model

aicc_values <- sapply(model_list, function(x) {
  AICc(x)
})

# Create comparison table

model_comparison <- data.frame(
  Model = names(model_list),
  AICc = aicc_values,
  Delta_AICc = aicc_values - min(aicc_values)
)

# Sort by AICc

model_comparison <- model_comparison[order(model_comparison$AICc),]

model_comparison

```

```{r top-model-probabilities, warning=FALSE}

# Get predictions without appendData to avoid row mismatch

occuProbs <- predict(m8, type = "state", na.rm = TRUE)
detProbs <- predict(m8, type = "det")

cat("Mean Occupancy Probability:", round(mean(occuProbs$Predicted),digits = 3), "\n") 
cat("Range of Occupancy Probability:", round(range(occuProbs$Predicted), digits = 3), "\n")

cat("Mean Detection Probability:", round(mean(detProbs$Predicted), digits = 3), "\n")
cat("Range in Detection Probability:", round(range(detProbs$Predicted), digits = 3), "\n")

```


## Interactive Map

```{r prep-data-for-map,, message=FALSE}

## Load and prep data

# Spatial Data

streams <- st_read("./Data/Spatial/Habitat/Water_features/WaterFeature_L.shp") # river shp

finalDF_ll <- df %>%
  select(Camera_Name, Assigned_Longitude, Assigned_Latitude)


# Site Info and Predictions

site_info <- m8@data@siteCovs

site_info <- site_info %>%
  filter(!(Camera_Name == "AFRL-NW-05" | 
              Camera_Name == "HQA-38-03")) # 2 cameras without data

occuProbsDF <- data.frame(
  site_info,
  OccuProb_Estimate = occuProbs$Predicted,
  SE = occuProbs$SE,
  CI_Lower = occuProbs$lower,
  CI_Upper = occuProbs$upper
   )

occuProbsDF <- left_join(occuProbsDF, finalDF_ll)

occuProbsDF <- occuProbsDF %>% # to export to Cole for report
  select(Camera_Name, Assigned_Longitude, Assigned_Latitude, OccuProb_Estimate, SE, CI_Lower, CI_Upper, Dist_to_Stream) %>%
  mutate(across(where(is.numeric), ~ round(., 3)))

write.csv(occuProbsDF, file = "C:/Users/akellner/OneDrive - Colostate/Documents/Edwards/Mojave Ground Squirrel/Outputs/Occupancy_Results_20250918.csv", row.names = FALSE)

```


```{r}

occuSF <- st_as_sf(occuProbsDF, coords = c("Assigned_Longitude", "Assigned_Latitude"), crs = "EPSG:4326") # create SF object

occUTM <- st_transform(occuSF, st_crs(streams)) # project to UTM 11N


# Map

tmap_mode('view')

occuMap <- tm_shape(streams) + 
  tm_lines(col = "#3D5B9F") +
tm_shape(occUTM) + 
  tm_symbols(fill = "OccuProb_Estimate",
             size = 0.5,
             popup.vars = c("Camera Name" = "Camera_Name",
                            "Occupancy Probability" = "OccuProb_Estimate",
                            "Distance to Stream" = "Dist_to_Stream" 
                            ),
             fill.scale = tm_scale_continuous(values = "viridis"),
             fill.legend = tm_legend(title = "Occupancy Probability", reverse = TRUE)) #+
  #tm_layout(legend.reverse = TRUE) 


tmap_save(occuMap, filename = "C:/Users/akellner/OneDrive - Colostate/Documents/Edwards/mgs_occupancy_map.html")

st_write(occUTM, "C:/Users/akellner/OneDrive - Colostate/Documents/Edwards/Mojave Ground Squirrel/Outputs/Occupancy_estimates_MGS_20250917.shp")




```

