---
title: "MGS_Occupancy_Final_Analysis"
author: "Annie Kellner"
date: "2025-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

```{r load-libraries, message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)
library(unmarked)
library(MuMIn)
library(sf)
library(tmap)
library(tmaptools)
library(viridisLite)

```


```{r load-data}

umf <- readRDS("./Data/UMFs/umf_allDateCols_06122025.Rds")

finalDF <- readRDS("./Data/indCams_adults_envCovs_06022025.Rds")

# Filter out cameras with no data BEFORE any analysis
finalDF <- finalDF %>%
  filter(!(Camera_Name == "AFRL-NW-05" | 
              Camera_Name == "HQA-38-03")) # 2 cameras without data

# Check if umf needs to be recreated with filtered data
# If umf was created with the full dataset, you may need to recreate it
# with the filtered finalDF to ensure consistency

```

## Model 1: Intercept-Only

```{r intercept-only, warning=FALSE}

m1 <- occu(~ 1 ~ 1,
           data = umf,
           linkPsi = "logit")

```


## Model 2
  - Detection trend only (no occupancy)
    - Linear date term only (scaled)

```{r det-trend-only-linear-scaled, warning=FALSE}

m2 <- occu(~ scaled_ordinal ~ 1,
           data = umf,
           linkPsi = "logit")
```

## Model 3
  - Detection trend only
    - Linear date term (recentered)

```{r det-trend-only-linear-recentered}

m3 <- occu(~ recentered_date ~ 1,
           data = umf,
           linkPsi = "logit")

```


## Model 4
  - Detection Trend Only
    - Linear + Quadratic (scaled)

```{r det-trend-only-linear-quadratic-scaled}

m4 <- occu(~ scaled_ordinal + scaled_ord_squared ~ 1,
           data = umf,
           linkPsi = "logit"
           )

```


## Model 5
  - Detection Trend Only
  - Linear + Quadratic (recentered)

```{r det-trend-linear-quadratic-recentered, warning=FALSE}

m5 <- occu(~ recentered_date + recentered_date_squared ~ 1,
           data = umf,
           linkPsi = "logit")

```


# AIC on Detection Trend

This AIC analysis is *only* looking at detection trends. It does not include any covariates for occupancy (i.e., environmental covariates)

Note: All models called "Quadratic" also include the linear term

```{r det-trend-AIC}

model_list <- list(
  "Intercept-Only" = m1,
  "Linear_Scaled" = m2,
  "Linear_Recentered" = m3,
  "Quadratic_Scaled" = m4,
  "Quadratic_Recentered" = m5
)

# Calculate AICc for each model

aicc_values <- sapply(model_list, function(x) {
  AICc(x)
})

# Create comparison table

model_comparison <- data.frame(
  Model = names(model_list),
  AICc = aicc_values,
  Delta_AICc = aicc_values - min(aicc_values)
)

# Sort by AICc

model_comparison <- model_comparison[order(model_comparison$AICc),]

model_comparison

```

It appears recentering the dates is the better approach. I am going to run another model selection on detection trend using recentered dates, but include a categorical variable for 'season' (early, peak, late). This is my final attempt to capture the parabolic detection trend.



## Model 6
  - Detection trend only
  - Linear Recentered Dates
  - Season

```{r warning=FALSE}

m6 <- occu(~ recentered_date + season ~ 1,
           data = umf,
           linkPsi = "logit")

```


## Model 7
  - Detection Trend Only
  - Quadratic Date Term (recentered)
  - Season Classification

```{r}

m7 <- occu(~ recentered_date + recentered_date_squared + season ~ 1,
           data = umf,
           linkPsi = "logit")

```

## Model 8
  - Season Only
  
```{r}

m8 <- occu(~ season ~ 1,
           data = umf, 
           linkPsi = "logit")
```
  

## AIC Analysis

```{r}

model_list <- list(
  "Intercept-Only" = m1,
  "Linear_Recentered" = m3,
  "Quadratic_Recentered" = m5,
  "Season Only" = m8,
  "Linear_Season" = m6,
  "Quadratic_Season" = m7,
  
  
)
 

# Calculate AICc for each model

aicc_values <- sapply(model_list, function(x) {
  AICc(x)
})

# Create comparison table

model_comparison <- data.frame(
  Model = names(model_list),
  AICc = aicc_values,
  Delta_AICc = aicc_values - min(aicc_values)
)

# Sort by AICc

model_comparison <- model_comparison[order(model_comparison$AICc),]

model_comparison

```

Dang!!! It worked!!!! 

## MODEL 9
  - Occupancy Covariate only
  
```{r occupancy-only}

m9 <- occu(~ 1 ~ Dist_Scaled,
           data = umf,
           linkPsi = "logit")

```


## MODEL 10
  - Detection covs: Quadratic trend on recentered dates + Season
  - Occupancy covs: Distance to Stream (scaled)
  
```{r quadratic-recentered-dist-to-stream}

m10 <- occu(~ recentered_date + recentered_date_squared + season ~ Dist_Scaled,
             data = umf,
             linkPsi = "logit")

```


## MODEL 11

```{r Quadratic-Distance}

m11 <- occu(~ recentered_date + recentered_date_squared ~ Dist_Scaled,
            data = umf,
            linkPsi = "logit")



```

```{r}
m12 <- occu(~ recentered_date ~ Dist_Scaled,
            data = umf, 
            linkPsi = "logit")
```


```{r}
m13 <- occu(~ season ~ Dist_Scaled,
            data = umf,
            linkPsi = "logit")
```


```{r}
m14 <- occu(~ recentered_date + season ~ Dist_Scaled,
            data = umf,
            linkPsi = "logit")
```


# AIC


```{r}

model_list <- list(
  "Intercept-Only" = m1,
  "Linear_Recentered" = m3,
  "Quadratic_Recentered" = m5,
  "Season Only" = m8,
  "Linear_Season" = m6,
  "Quadratic_Season" = m7,
  "Distance-Only" = m9,
  "Linear_Distance" = m12,
  "Quad_Distance" = m11,
  "Quad_Season_Distance" = m10,
  "Season_Distance" = m13,
  "Linear_Season" = m14
)


# Calculate AICc for each model

aicc_values <- sapply(model_list, function(x) {
  AICc(x)
})

# Create comparison table

model_comparison <- data.frame(
  Model = names(model_list),
  AICc = aicc_values,
  Delta_AICc = aicc_values - min(aicc_values)
)

# Sort by AICc

model_comparison <- model_comparison[order(model_comparison$AICc),]

model_comparison

```

The following are means and ranges for occupancy and detection probabilities across all sites and throughout the study duration. 

```{r top-model-probabilities, warning=FALSE}

# Get predictions without appendData to avoid row mismatch
occuProbs <- predict(m10, type = "state", na.rm = TRUE)
detProbs <- predict(m10, type = "det")

cat("Mean Occupancy Probability:", round(mean(occuProbs$Predicted),digits = 3), "\n") 
cat("Range of Occupancy Probability:", round(range(occuProbs$Predicted), digits = 3), "\n")

cat("Mean Detection Probability:", round(mean(detProbs$Predicted), digits = 3), "\n")
cat("Range in Detection Probability:", round(range(detProbs$Predicted), digits = 3), "\n")

```

## Detection Probability Over Time

```{r plot-detection-prob}

# Create dataframe with detProbs and dates

detProbDF <- data.frame(
  "Date" = colnames(finalDF[,2:111]),
  "Detection_Prob" = detProbs$Predicted
)

detProbDF$Date <- mdy(detProbDF$Date)

ggplot(detProbDF, aes(x = Date, y = Detection_Prob)) + 
  #geom_point() + 
  geom_line() + 
  scale_x_date(date_breaks = "15 days",
               date_labels = "%m/%d") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Date", y = "Detection Probability")


```

## Actual captures over time

```{r plot-captures-over-time}

counts <- finalDF %>%
  summarize(across(2:111, sum, na.rm = TRUE))

counts_long <- counts %>%
  pivot_longer(everything(),
               names_to = "Date",
               values_to = "Total_Detections") 

counts_long <- counts_long %>% mutate(Date = mdy(Date))


# Plot Count Data

ggplot(counts_long, aes(x = Date, y = Total_Detections)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Date",
       y = "Number of Detections",
       title = "Daily Detection Counts") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_date(date_breaks = "5 days",  # Show every 5 days
               date_labels = "%m/%d")    # Format as month/day


```



## Interactive Map

```{r prep-data-for-map,, message=FALSE}

## Load and prep data

# Spatial Data

streams <- st_read("./Data/Spatial/Habitat/Water_features/WaterFeature_L.shp") # river shp

finalDF_ll <- finalDF %>%
  select(Camera_Name, Assigned_Longitude, Assigned_Latitude)


# Site Info and Predictions

site_info <- m10@data@siteCovs

site_info <- site_info %>%
  filter(!(Camera_Name == "AFRL-NW-05" | 
              Camera_Name == "HQA-38-03")) # 2 cameras without data

occuProbsDF <- data.frame(
  site_info,
  OccuProb_Estimate = occuProbs$Predicted,
  SE = occuProbs$SE,
  CI_Lower = occuProbs$lower,
  CI_Upper = occuProbs$upper
   )

occuProbsDF <- left_join(occuProbsDF, finalDF_ll)

occuProbsDF <- occuProbsDF %>% # to export to Cole for report
  select(Camera_Name, Assigned_Longitude, Assigned_Latitude, OccuProb_Estimate, SE, CI_Lower, CI_Upper, Dist_to_Stream) %>%
  mutate(across(where(is.numeric), ~ round(., 3)))

write.csv(occuProbsDF, file = "C:/Users/akellner/OneDrive - Colostate/Documents/Edwards/Mojave Ground Squirrel/Outputs/Occupancy_Results_09112025.csv", row.names = FALSE)

```


```{r map}

occuSF <- st_as_sf(occuProbsDF, coords = c("Assigned_Longitude", "Assigned_Latitude"), crs = "EPSG:4326") # create SF object

occUTM <- st_transform(occuSF, st_crs(streams)) # project to UTM 11N


# Map

tmap_mode('view')

occuMap <- tm_shape(streams) + 
  tm_lines(col = "#3D5B9F") +
tm_shape(occUTM) + 
  tm_symbols(fill = "OccuProb_Estimate",
             size = 0.5,
             popup.vars = c("Distance to Stream" = "Dist_to_Stream", 
                            "Camera Name" = "Camera_Name"),
             fill.scale = tm_scale_continuous(values = "viridis"),
             fill.legend = tm_legend(title = "Occupancy Probability", reverse = TRUE)) #+
  #tm_layout(legend.reverse = TRUE) 


#tmap_save(occuMap, filename = "C:/Users/akellner/OneDrive - Colostate/Documents/Edwards/mgs_occupancy_map.html")

#st_write(occUTM, "C:/Users/akellner/OneDrive - Colostate/Documents/Edwards/Mojave Ground Squirrel/Outputs/Occupancy_estimates_MGS_09112025.shp")



```

